---
title: "Forest Fires in Portugal"
author: João Teixeira - up200705307, Nuno Peixoto - up200804621
date: "25-01-2015"
output:
  beamer_presentation:
    toc: true
theme: Antibes
---

```{r imports, echo=FALSE, results='hide', comment=FALSE, message=FALSE}
setwd(dir = "/home/joao/Documents/Data-Mining")
source('Functions.R')
train_data <- read.csv("data/train.csv", head=T,  na.strings=c("?"))
```


## Introduction

* Monitoring and forecasting forest fires in Portugal;

* The several variables may influence the burnt area;

* In 2003, Portugal faced the worst forest fire losing 8.6% of of the total area;

* Elevation, slope or density are some of the specifications of the data set; 

**Objective** : Explore and predicte the data of the forest.


## Exploratory analysis of the data

* Global Summary
* Main Variables
* Target Variable

## Global Summary

* Number of Columns: `r ncol(train_data)`.
* Number of Rows: `r nrow(train_data)`.
* Number of Data: `r ncol(train_data)*nrow(train_data)`.
* Target Value: 1 (TotalBurntArea) - Numeric variable.
* Number of Unknown Values: `r sum(apply(train_data, 1, function(x){any(is.na(x))}))`.

## Global Summary (cont.)

**Climate Variables** - The climatic conditions may affect the probability of a fire to occur;

**Landscape Variables** - The landscape has been extensively associated with fire occurrence;

**Socio-economic Variables** - Human have impact in historical fire patterns;

**Topographic Variables** - The topographic features may influence the fire ignitions;


## Main Variables 

In the following table we have the **TOP5** main variables: 

```{r gain_ratio, echo=FALSE}
a <- gain.ratio(formula =  TotalBurntArea ~ ., train_data)
a$attribute <- rownames(a)
a <- head(a[order(a$attr_importance, decreasing =T ),],5)
rownames(a) <- 1:nrow(a)
pander(a, style = "grid")
```

## Main Variables (Number of outliers)
```{r top5_vars, echo=FALSE, comment=NA}
att <- a$attribute
```

```{r call_functions, echo=FALSE, results='hide'}
nouts <- out_tax(train_data,att)
stand_dev <- stand_dev(train_data,att)
```

* ELEV_MAX: `r nouts[1]`

* Bio1: `r nouts[2]`

* ELEV_MEAN: `r nouts[3]`

* Bio7: `r nouts[4]`

* DensPop01: `r nouts[5]`


## Main Variables (Standard Deviation)
* ELEV_MAX: `r stand_dev[1]`

* Bio1: `r stand_dev[2]`

* ELEV_MEAN: `r stand_dev[3]`

* Bio7: `r stand_dev[4]`

* DensPop01: `r stand_dev[5]`

## Target Variable

```{r tv_sum, echo=FALSE, comment=NA}
tmp <-summary(train_data["TotalBurntArea"])
prc <- summaryToDataFrame(tmp)
pander(prc, style="grid")
```

## Target Variable (Number of outliers)
```{r tv_out, echo=FALSE}
outs <- out_tax(train_data,c("TotalBurntArea"))
```

* TotalBurntArea: `r outs[1]`

We can see that more than 10% of the total burnt area values are considered outliers.

## Target Variable (Total Area vs. Total Burnt Area)
```{r times_burnt, echo=FALSE}
tmp<-train_data[train_data$TA<train_data$TotalBurntArea,c("TA","TotalBurntArea")]
ggplot(tmp, aes(factor(round(tmp$TotalBurntArea/tmp$TA)))) + geom_bar() + xlab("Times Burnt") + ggtitle("Times Total Area Burnt")
```


## Data Pre-Processing

* Remove None importance Variables
* Normalizing Value

## Remove None importance Variables

```{r non_importance_vars, echo=FALSE}
nanration <- gain.ratio(formula =  TotalBurntArea ~ ., train_data)
nanration$attribute <- rownames(nanration)
nanration <- nanration[order(nanration$attr_importance, decreasing =T ),]
nanration <- nanration[is.nan(nanration$attr_importance),]
rownames(nanration) <- 1:nrow(nanration)
pander(nanration, style = "grid")
```


```{r remove_ration, echo=FALSE, results='hide'}
data <- ncol(train_data)*nrow(train_data)
rmdata <- nrow(train_data)*(ncol(train_data)-nrow(nanration))
perc <- round(((data-rmdata)/(data))*100,2)
```

* This pre-processing that does not prejudice the information analysis;
* We reduce about `r perc`% of the total data.

## Normalizing Value

* Data normalization pre-processing we will use for the analisys;

```{r normalize, echo=FALSE}
norm.train_data <- scale(train_data[,])
colnames(train_data) <- colnames(train_data)
norm.train_data <- data.frame(norm.train_data)
```

```{r view_norm_data, echo=FALSE}
head(train_data[,c(1:6)], 3)
head(norm.train_data[,c(1:6)], 3)
```

